\vspace{-0.25em}\thetitle

### 2009-07-24 paper.tex
...

A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
A 1500-dimensional vector can be used to count the frequency of packet sizes seen, from one byte up to the typical \caps{MTU} of 1500 bytes.
In this representation, $\Phi(s)$ has as its $i$th entry the number of packets in $s$ of size $i$.
Since \caps{IP} protocols range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
Histogram representations for port numbers and inter-packet intervals are detailed in \Section{methodology}.
% Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
% For practical purposes, however, this is too large a vector space, so we represent port numbers in 1000-dimensions, and reduce port numbers modulo 1000.
% Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
The representation we use is a direct sum of these histogram representations.
The combination is a representation in 5256 dimensions.
An entire trace is represented as a matrix of such representation vectors---one flow per row.
This paper explores the linear structure of such matrices.

@@

To see the linear structure of these representation matrices, we apply one of the most powerful tools of linear algebra:
the singular value decomposition (\caps{SVD}).
This matrix decomposition can be applied to any real matrix, factoring it into a product of three real matrices:
\begin{align}
  X = U\!S\trans{V}.
\end{align}
$U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
Among the useful properties of this factorization, is that the first $k$ columns of $U\!S$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
This property allows us to visualize the structure of $X$ using initial columns of $U\!S$.
However, we are interested only interested in direction since
vectors differing only by magnitude lie in the same one-dimensional subspace.
To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane from the origin.\projectionnote

@@

\section{Whence Linearity?}

Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be posed in the form of a model.
% A mixture model, to be specific.
We propose viewing each flow as having a probability distribution of possible behaviors it could exhibit.
% We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
Moreover, only even smaller subsets of these basic behaviors are ever mixed with each other.
Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
\begin{align}\eqn{mixture-model}
  q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
\end{align}
Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.

>>>>

A particularly powerful class of linear representations describes the behavior of collections of flows using histogram vectors to count the frequencies of various features.
A 1500-dimensional vector can be used to count the frequency of packet sizes seen, from one byte up to the typical \caps{MTU} of 1500 bytes.
In this representation, $\Phi(s)$ has as its $i$th entry the number of packets in $s$ of size $i$.
Since \caps{IP} protocols range from 0 to 255, the frequency of types in a set of flows can be represented in 256 dimensions.
Histogram representations for port numbers and inter-packet intervals are detailed in \Section{methodology}.
% Source and destination port numbers can similarly be represented using 65536-dimensional vectors.
% For practical purposes, however, this is too large a vector space, so we represent port numbers in 1000-dimensions, and reduce port numbers modulo 1000.
% Finally, we represent inter-packet intervals by first quantizing into 1500 bins of interval values as described by Karpinski~\emph{et~al.} and counting the number of intervals in flows occurring in each bin.
The representation we use is a direct sum of these histogram representations.
% The combination is a representation in 5256 dimensions.
An entire trace is represented as a matrix of such representation vectors---one flow per row.
This paper explores the linear structure of such matrices.

@@

To see the linear structure of these representation matrices, we apply one of the most powerful tools of linear algebra:
the singular value decomposition (\caps{SVD}).
This matrix decomposition can be applied to any real matrix, factoring it into a product of three real matrices:
\begin{align}
  X = US\trans{V}.
\end{align}
$U$ and $V$ are orthogonal and $S$ is diagonal, with entries of nonincreasing nonnegative value.
Among the useful properties of this factorization, is that the first $k$ columns of $US$ are an optimal $k$-dimensional reduced model of $X$.\svdnote
This property allows us to visualize the structure of $X$ using initial columns of $US$.
However, we are interested only interested in direction since
vectors differing only by magnitude lie in the same one-dimensional subspace.
To visualize direction while eliminating distance, we project all data points onto the unit-sum hyperplane from the origin.\projectionnote

@@

\section{Model \& Hypothesis}

Our answer to the question of why low-dimensional linear structure appears in the feature-frequency representation of network traffic can be expressed as a stochastic model of flow behavior.
% A mixture model, to be specific.
We propose viewing each flow as having a probability distribution of possible behaviors it could exhibit.
% We propose that the behavior of each flow is drawn from a probability distribution on a space possible behaviors.
We hypothesize that the behavior distribution for most flows is a mixture of a relatively small set of ``basic behaviors.''
Moreover, only even smaller subsets of these basic behaviors are typically combined with each other.
Under these assumptions, we can express the distribution of each flow's behaviors as a finite mixture model~\cite{McLachlan00}:
\begin{align}\eqn{mixture-model}
  q_i(x) = \sum_{j=1}^r w_{ij} p_j(x),
\end{align}
Here $q_i$ and $p_j$ are probability density functions, and $w_{ij}$ are nonnegative weights, summing to unity for each $i$.
\Equation{mixture-model} is expressed succinctly as matrix multiplication.
Writing $Q_{ik} = q_i(k)$, $W_{ij} = w_{ij}$, and $P_{jk} = p_j(k)$, we have:
\begin{align}\eqn{mixture-model-matrix}
  Q = WP.
\end{align}
The number of basic behaviors, $r$, is the maximum possible rank of the feature distribution matrix, $Q$.
Next, partition the rows of $P$ into small sets which can have simultaneously non-zero weights: $P = [P_1;\cdots;P_g]$ using Matlab concatenation notation;
assume the rows of $P$ are ordered to allow this.
We can also assume that the rows of $Q$ and $W$ are ordered so that we can write $W = W_1 \directsum \cdots \directsum W_g$.
